---
title: "Final Project"
output: html_notebook
authors: "Joshua Allessio, Kashyap Vallur and Marco Camalich"
---

## Problem Statement
Washington DC's public bikeshare dataset is a classic data science project. Much can be gleaned from this data. In this report, we set out to discover which of the bike stations are particularly efficient or inefficient. The goal is to create a visualization of the net flow of bikes for each station over time. 

If the net flow for a bike station often stays close to 0, we may assume that that station is in an optimal position and hosts enough bicycles. Furthermore, high turnover of bikes can imply that a bike station is in an optimal position, while low turnover of bikes may imply that that bike station is not placed in an optimal location. 

Another pattern that may arise could be a bike station that has a strong negative netflow early in the day, followed by a stagnant period of netflow later in the day. This would imply that the bike station has quickly become vacant - now offering no bicycles for potential riders. Potential solution for this particular problem might be increasing the size of that station, creating a new one nearby, or making a larger change to the infrastructure of the bikeshare. 

## Document Setup
Install packages required for analysis.
```{r}
required_packages <- c("tidyverse", "janitor", "here", "openmeteo", "lubridate", "forcats", "dplyr")

for (packages in required_packages) {
  if (!require(packages, character.only = TRUE)) {
    install.packages(packages, dependencies = TRUE)
    library(packages, character.only = TRUE)
  }
}

if (all(sapply(required_packages, requireNamespace, quietly = TRUE))) {
  cat("All required packages are loaded.\n")
} else {
  warning("Some packages failed to load.")
}

```

## Create a function to filter useless data from the data set
This function takes the DC bikeshare data sets and removes columns with NA values, as well as rides which lasted
less than 30 seconds or more than 13 hours. 
```{r}
filter_bikeshare <- function(df) {
  df |> filter(
    !is.na(start_station_name) & !is.na(start_station_id) & start_station_name != "" & start_station_id != "" &
    !is.na(end_station_name) & !is.na(end_station_id) & end_station_name != "" & end_station_id != "" &
    !is.na(start_lat) & !is.na(start_lng) & !is.na(end_lng)    
  )
  df <- df |> 
    mutate(duration = as.numeric(difftime(ended_at, started_at))) |>  
    filter(duration > 30) |>  # Longer than 30 seconds
    filter (duration < 46800) # Less than 13 hours
return(df)
}
```


## Import the data
```{r}
df_01=read_csv(here("data_raw", "202301-capitalbikeshare-tripdata.csv"))
df_02=read_csv(here("data_raw", "202302-capitalbikeshare-tripdata.csv"))
df_03=read_csv(here("data_raw", "202303-capitalbikeshare-tripdata.csv"))
df_04=read_csv(here("data_raw", "202304-capitalbikeshare-tripdata.csv"))
df_05=read_csv(here("data_raw", "202305-capitalbikeshare-tripdata.csv"))
df_06=read_csv(here("data_raw", "202306-capitalbikeshare-tripdata.csv"))
df_07=read_csv(here("data_raw", "202307-capitalbikeshare-tripdata.csv"))
df_08=read_csv(here("data_raw", "202308-capitalbikeshare-tripdata.csv"))
df_09=read_csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
df_10=read_csv(here("data_raw", "202310-capitalbikeshare-tripdata.csv"))
```

## Tidy the Data
```{r}
df_01 <- filter_bikeshare(df_01)
df_02 <- filter_bikeshare(df_02)
df_03 <- filter_bikeshare(df_03)
df_04 <- filter_bikeshare(df_04)
df_05 <- filter_bikeshare(df_05)
df_06 <- filter_bikeshare(df_06)
df_07 <- filter_bikeshare(df_07)
df_08 <- filter_bikeshare(df_08)
df_09 <- filter_bikeshare(df_09)
df_10 <- filter_bikeshare(df_10)
```

## Aggregate the Data
Aggregate the data into one data frame and begin with a simple visualization
```{r}
all_data <- bind_rows(
  df_01, df_02, df_03, df_04, df_05, df_06, df_07, df_08, df_09, df_10,
  .id = "month"
)

all_data$started_at <- as.POSIXct(all_data$started_at, format = "%Y-%m-%d %H:%M:%S")
all_data$month <- format(all_data$started_at, "%Y-%m")

ggplot(all_data, aes(x = month)) +
  geom_bar() +
  labs(title = "Monthly Ride Counts",
       x = "Month",
       y = "Ride Counts") +
  theme_minimal()


```

## Find out how many bike stations there are
```{r}
unique_station_count <- all_data %>%
  filter(!is.na(start_station_name) & !is.na(start_station_id) & start_station_name != "" & start_station_id != "") %>%
  summarise(unique_stations = n_distinct(start_station_name))

cat("The total amount of unique stations is:", unique_station_count$unique_stations, "\n")
```

## Measure Activity At Each Station
In this code chunk, we count how many times each station name appears in the data set. The more a station's name appears, the more it is used. We will be able to see which stations are the most useful, and which stations not being used in any significant capacity. 
```{r}
# Combine start and end stations into one column
all_stations <- bind_rows(
  select(all_data, station = start_station_name),
  select(all_data, station = end_station_name)
)

# Get the top 5 stations (excluding NA)
top_stations <- all_stations %>%
  filter(!is.na(station)) %>%
  count(station, sort = TRUE) %>%
  head(5)

print(top_stations)
```
Above, we can see the stations that were used the most during the period of observation. Columbus Circle bike station was used 78000 times from January to November 2023. 

## Create a visualization
```{r}
# Convert month to a factor with the correct order
all_data$month <- factor(all_data$month, levels = unique(all_data$month))

# Filter data for the top 5 stations by activity
top_stations_data <- all_data %>%
  filter(start_station_name %in% top_stations$station)

# Create the line plot with increased line width (e.g., size = 1.5)
ggplot(top_stations_data, aes(x = month, color = start_station_name, group = start_station_name)) +
  geom_line(stat = "count", size = 1) +  # Adjust the size value as needed
  scale_x_discrete(name = "Month") +
  scale_y_continuous(name = "Ride Counts") +
  labs(title = "Monthly Ride Counts for Top 5 Stations",
       color = "Station") +
  theme_minimal() +
  theme(legend.position = "top")
```
In the visualization above, we can see the monthly change in bike station activity for the 5 most used bike stations. We can see that the Lincoln Memorial station is used much more in November than it was in February. 

## Continue Observing Bike Station Activity
Now, we will see which stations were used the least. 
```{r}
# Get the least active stations
worst_stations <- all_stations %>%
  filter(!is.na(station)) %>%
  count(station, sort = TRUE) %>%
  filter(n < 50) %>%
print(worst_stations)


# Get 5 stations with little activity
bad_stations <- all_stations %>%
  filter(!is.na(station)) %>%       # exclude NA values
  count(station, sort = TRUE) %>%   # count how many times each station was used
  filter(n >= 50) %>%               # exclude 
  tail(5)
print(bad_stations)
```
As you can see, the least used station is located at Shady Grove Hospital, with only 2 uses this year. That's a big difference between the most used station, at 78,000 uses this year! Upon further research, we can discove that Shady Grove Hospital has been permanently closed. This news, in combination with the fact that the bike station was used only twice, makes a good case for the removal of this particular station. It should be removed, or relocated somewhere where it will be more useful. 



```{r}
ggplot(all_data, aes(x = month, fill = rideable_type)) +
  geom_bar(position = "fill") +
  labs(title = "Popular Rideable Types by Month",
       x = "Month",
       y = "Proportion") +
  theme_minimal() +
  scale_y_continuous(labels = scales::percent_format(scale = 100))
```


