---
title: "Bike station deficits/gains"
output: html_notebook
---

Marco Camalich, Joshua Alessio and Kashyap Vallur

```{r}
library(tidyverse)
library(janitor)
library(here)
library(openmeteo)
library(lubridate)
library(forcats)
```

```{r}
df1=read_csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
```
```{r}
dfb <- df1 %>%
  filter(
    ymd_hms(started_at) <= ymd_hms(ended_at),
    !is.na(start_station_id),
    !is.na(end_station_id)
  )
```

```{r}
dfb_starts <- dfb %>%
  group_by(start_station_id) %>%
  summarise(starts = n())
dfb_starts
```

```{r}
dfb_ends <- dfb %>%
  group_by(end_station_id) %>%
  summarise(ends = n())
dfb_ends
```

```{r}
dfb_join <- left_join(dfb_starts, dfb_ends, by = c("start_station_id" = "end_station_id"))
dfb_join$station_id <- dfb_join$start_station_id
dfb_join$start_station_id <- NULL
dfb_join <- dfb_join %>%
select(station_id, everything())
dfb_join
```
```{r}
dfb_join <- dfb_join %>% mutate(net_gain = starts - ends)
dfb_join
```
```{r}
dfb_join$station_id <- as.factor(dfb_join$station_id)
summary(dfb_join)
```

```{r}
ggplot(dfb_join, aes(x = as.factor(station_id), y = net_gain, fill = factor(net_gain > 0))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("red", "blue"), guide = FALSE) +
  labs(title = "Net Gain by Station",
       x = "Station ID",
       y = "Net Gain") +
  theme_minimal() +
  coord_cartesian(ylim = c(-50, 50))  # Adjust the limits based on your data
```
```{r}
top_bottom_stations <- dfb_join %>%
  arrange(net_gain) %>%
  slice_head(n = 10) %>%
  bind_rows(dfb_join %>%
              arrange(desc(net_gain)) %>%
              slice_head(n = 10))

ggplot(top_bottom_stations, aes(x = fct_reorder(factor(station_id), net_gain), y = net_gain, fill = factor(net_gain > 0))) +
  geom_bar(stat = "identity", alpha = 0.7) +
  scale_fill_manual(values = c("red", "blue"), guide = FALSE) +
  labs(title = "Top and Bottom 10 Stations by Net Gain",
       x = "Station ID",
       y = "Net Gain") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
*In conclusion, we can observe that Station 31215 is the one that has the most ends of trips while 31602 is the one used for more starts.*

**Hypotheses:**

- Are there significant differences in usage levels among bike stations?

- Do variations exist in the average net gain of bikes during different times of the day?

- Are there differences in bike usage patterns between weekdays and weekends?

- Does net gain vary significantly between members and casual users?

- Are there significant differences in bike net gain based on geographical locations?




