---
title: "R Notebook"
output: html_notebook
---
Load Packages
```{r} 
library(tidyverse)
library(janitor)
library(here)
library(openmeteo)
library(lubridate)
```

```{r}
df1=read_csv(here("data_raw", "202309-capitalbikeshare-tripdata.csv"))
```


```{r}
df2s=df1 %>% 
  select(rideable_type,member_casual,
                    contains("start"),ride_id) %>% 
  mutate(start_stop="start") %>%
  rename(t=started_at,
         station_name=start_station_name,
         station_id=start_station_id,
         lat=start_lat,
         lng=start_lng)
```


```{r}
df2e=df1 %>% 
  select(ride_id,rideable_type,member_casual,
                    contains("end")) %>%
  mutate(start_stop="stop") %>%
  rename(t=ended_at,
         station_name=end_station_name,
         station_id=end_station_id,
         lat=end_lat,
         lng=end_lng)
```

```{r}
df2=bind_rows(df2s,df2e) %>%
  arrange(t) %>%
  mutate(rider_delta=(start_stop=="start")*2-1) %>% #change in ridership 
  mutate(riders=cumsum(rider_delta)) %>%
  relocate(riders,.after=t)
```

```{r}
df2 %>% 
  ggplot(aes(t,riders)) +
  geom_line()
```

# Constructing a subsampled dataset
Let's try to create a subsample of the dataset so our workflow doesn't have to run through 450,000 calculations each time we test the code
```{r}
df_s <- df2 %>% slice_head(n=1000)        # Take off the top 1000 rows of df2
```


Use a smaller dataset to check to see if our process is correct before we perform it on the entire huge dataset
```{r}

df_e=df_s |>
  mutate(t_f=floor_date(t,"10 mins")) %>%  # Add new column, which rounds down the time to nearest 10 minutes
  relocate(t_f,.after=t) %>%               # rearrange column order - put "t_f" to the right of "t"
  slice_head(n=1, by=t_f)                  # Take the first row of each different value of t_f by treating t_f as a                                                  grouping variable 
df_e
```
```{r}
df_r=df2 |>
  mutate(t_f=floor_date(t,"1 mins")) %>%
  relocate(t_f,.after=t) %>%
  slice_head(n=1,by=t_f)
```


Visualize the difference between the full dataset and the sample dataset
```{r}
p1=df2 %>% 
  filter(day(t)==18) %>%
  ggplot(aes(t,riders)) +
  geom_line() +
  ggtitle("Riders on 18Sep")
p1+
  geom_line(data=df_r %>% filter(day(t)==18),
  color="red")                              
```



# Get Weather Data for September
Using the openmeteo package
```{r}
df_w=weather_history("Washington",
                    start = "2023-09-01",
                    end = "2023-09-30",
                    hourly = c("apparent_temperature",
                               "wind_speed_10m",
                               "precipitation")
)
```




This code block should reveal a previous mistake - the time zones are different
```{r}
df2c=df2 %>% mutate(t=force_tz(t,tzone="America/New_York")) #corrected
 
df_s2=df2c %>% slice_sample(n=1000)
 
df_j2=df_s2 %>% 
  left_join(df_w,by=join_by(closest(t>=datetime)))  %>%
  relocate(datetime, .after=t)
 
head(df_j2)
```


```{r}

df2$t[1:5]
force_tz(df2$t[1:5],"America/New_York")
    df2c=df2 %>% mutate(t=force_tz(t,tzone="America/New_York")) #corrected
df_s2=df2c %>% slice_sample(n=1000)
df_j2=df_s2 %>% 
  left_join(df_w,by=join_by(closest(t>=datetime)))  %>%
  relocate(datetime, .after=t)
head(df_j2)

```


```{r}
df1d = df1 %>%
  mutate(ride_time=ended_at - started_at) %>%
df1d <- 
```

```{r}
df1d %>% 
  ggplot(aes(ride_time)) +
  geom_histogram()
```
